Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study

### 2021-06-22 submission/manuscript.rmd
reduce abstract word count

---
title: "Smooth risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
abstract: |
  **Objective:** Compare different risk-based approaches to estimating
  individualized treatment effects with simulations of the RCT setting. **Study
  Design and Setting:** We compared different methods for predicting absolute
  benefit on the basis of an available prognostic index (PI) for risk: A model
  with the PI and a constant relative treatment effect; models including an
  interaction of treatment assignment with the PI; 4 quarters of the PI; nonlinear
  transformations of the PI with increasing flexibility (restricted cubic splines
  with 3, 4 and 5 knots, respectively); an adaptive method, selecting the optimal
  model based on Akaike’s Information Criterion. Starting from a base case
  scenario (sample size 4,250, average treatment effect odds ratio 0.8, AUC of the
  PI 0.75), we considered diverse assumptions of the data generation process of
  the “true” treatment effect, introducing linear and quadratic interactions of
  the PI with treatment. We analyzed the effect of varying sample size and
  discriminative ability of the PI. We evaluated performance using root mean
  squared error, discrimination for benefit and calibration for benefit. **Results:**
  The model including a linear interaction of the risk linear predictor with
  treatment had adequate performance that was robust under the majority of the
  simulation scenarios. Methods using restricted cubic splines required larger
  sample sizes and higher AUC of the PI to achieve adequate performance. The
  adaptive approach’s performance was comparable to the performance of the best
  model in each scenario. **Conclusion:** In most cases using a model just
  including a linear interaction of the PI with treatment
  adequately predicts absolute benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - test
  - test that
output: 
  rticles::arxiv_article:
    keep_tex: true
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

In the simulated datasets based treatment was allocated at random using a 50/50
split. For each patient we simulated $8$ baseline covariates, where
$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes
for patients in the control arm were generated from a logistic regression model
including all baseline covariates. Coefficient values were such, so that the
prediction model had an AUC of $0.75$ and an event rate of $20%$ in the control
arm was achieved. Under the base case scenario, outcomes in the treatment arm
were created using the same logistic regression model, including a constant
treatment odds ratio (OR) of $0.8$. The generated samples of the base case
scenario were 4,250 ($80\%$ power for the detection of an unadjusted OR of
$0.8$).

We evaluated the effect of sample size considering additional scenarios with
sample sizes of `r 1064` and `r 17000`. We also evaluated the effect of prediction
performance, adjusting the baseline covariate coefficients, so that AUC values
of $0.65$ and $0.80$ were achieved when validating in a simulated dataset of
$500,000$ patients.

We simulated binary outcomes in the control arm based on 
$\text{expit}(PI)$, where 
$\text{expit}(x)= \frac{e^x}{1 + e^x}$. 
In the treatment arm outcomes where generated from $\text{expit}(lp_1)$ with
$$ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,$$
where the coefficients $\gamma_0, \gamma_1$ and $\gamma_2$ along with the
centering constant $c$ were set for each simulation scenario. In this way we
were able to assess a wide variety of scenarios, ranging from true constant
relative treatment effects to moderate and strong linear and quadratic
deviations. We also considered scenarios with treatment-covariate interactions.
These scenarios include 4 weak interactions
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
4 strong interactions 
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), 
and 2 weak and 2 strong interactions. Combining all these different settings
resulted in a simulation study of $66$ scenarios. The exact settings for each
scenario are available in the supplementary material.

@@

All methods assume that a risk prediction model is available and can be used to
assign individualized predictions. For the simulations we developed the
prediction models internally and blinded to treatment using logistic regression
including main effects for all baseline covariates and treatment. Predictions on
individuals were made setting treatment to $0$.

The **stratified HTE method** was suggested as an alternative to traditional
subgroup analyses. Patients are stratified into equally-sized risk strata---in
this case based on risk quartiles. Absolute effects are estimated using the
differences in event rates between treatments within risk quarters. We
considered this approach as a reference, expecting it to perform worse than the
other candidates, as its objective is not individualized benefit prediction.

We also considered a set of **linear methods**. We fit separate models within
treatment arms using only the treatment indicator and the linear predictor of
the internal risk prediction model. In the simpler case, we assume a constant
relative treatment effect (OR). Absolute benefit is then estimated from
$\text{expit}(lp +\log(\text{OR}))$, 
where $\text{expit}(x)=\frac{e^x}{1+e^x}$
and *lp* is the linear predictor of the prediction model. A different approach
fits a logistic regression using treatment, risk linear predictor and their
interaction. In this case, absolute benefit is estimated from
$\text{expit}(\beta_0+\beta_{lp}lp) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{lp}+\beta_*)lp)$.
We will refer to this method as the linear interaction approach.

Finally, we used restricted cubic splines (RCS) to relax the linearity assumption on
the effect of the linear predictor [@Harrell1988]. We compared the results
for 3, 4 and 5 knots when fitting the splines to introduce increasing
flexibility to the methods considered.

@@

Assuming that $\tau(\bm{x})=E\{y\given\bm{x}, t_x=0\} - E\{y\given\bm{x},t_x=1\}$ is the true benefit for each patient and $\hat{\tau}(x)$ is the estimated benefit from a method under study, the ideal loss function to use for the considered methods would be the unobservable root mean squared error $E\big\{(\hat{\tau} - \tau)^2\given \bm{x}\big\}$. However, in simulations the true patient-level benefit is available. Therefore, we estimated RMSE from
$$\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2$$

We also compared the predictive performance of the methods under study. We
assessed discrimination using the c-for-benefit statistic described in
[@vanKlaveren2018]. Patients in each treatment arm are ranked based on their
predicted benefit and then are matched 1:1, dropping patients in the larger
treatment arm without a pair. We define *observed* pair-specific treatment
benefit as the difference of observed outcomes between the untreated and the
treated patient of each pair. Pair-specifi *predicted* benefit is defined as
the average of predicted benefits within each pair. Then, c-for-benefit is
defined as the probability that from two randomly chosen matched patient pairs
with unequal observed benefit, the pair with greater observed benefit also has a
higher predicted benefit.

We evaluated calibration in a similar manner, using the integrated calibration
index (ICI) for benefit [@Austin2019]. After creating pairs based on predicted
benefit, observed benefits are regressed on the predicted benefits using a
locally weighted scatterplot smoother (loess). The ICI is the area between the
loess fit and the diagonal, with values closer to $0$ indicating better
calibration.

@@

The model including a constant relative treatment effect had the lowest median
RMSE in scenarios where the base case of true constant relative treatment effect
(OR = 0.8, N = `r 4250` and AUC = 0.75) or moderate relative deviations were
considered (Figure \ref{fig:rmsebase}; Panel A). However, when we considered
strong linear and quadratic deviations from the base case the linear interaction
model performed best (Figure \ref{fig:rmsebase}; Panels B and C). Only in the
case of strong quadratic deviations models including RCS smoothing (3 knots)
performed equally well to the linear interaction method. Increasing the number
of knots in RCS smoothing resulted in higher error rates across all scenarios.
The performance of the adaptive approach was quite comparable with the best
performing model in each scenario.

@@

When we increased the sample size (N = `r 17000`), in all scenarios the model more
in agreement with the underlying settings had the lowest error rates. Under the
base case of constant relative treatment effects (OR = 0.8), the model including
a constant relative treatment effect had the lowest RMSE (Figure
\ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong linear
deviations the linear interaction model performed best (Figure
\ref{fig:rmsesamplesize}; Panels B and C). In the case of true strong quadratic
deviations from the base case of constant relative effects, the more flexible
RCS smoothing models (3 knots) had the lowest RMSE. Again, the increased
flexibility of RCS smoothing with higher number of knots resulted in overfitting
and worse performance (Figure \ref{fig:rmsesamplesize}; Panel D).

>>>>

---
title: "Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
abstract: |
  **Objective:** Compare risk-based methods for individualizing treatment 
  effects with simulations in the RCT setting. **Study Design and Setting:** We
  predicted absolute benefit based on an available prognostic index (PI) using:
  a model with the PI and a constant relative treatment effect; a model 
  including an interaction of treatment with the PI; 4 quarters of the PI; 
  nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5
  knots); an adaptive model selection method using Akaike’s Information 
  Criterion. Starting from a base case scenario (sample size 4,250, constant 
  odds ratio 0.8, AUC of the PI 0.75), we considered diverse assumptions by
  introducing linear and quadratic interactions of the PI with treatment and 
  varying sample size and discriminative ability of the PI. We evaluated
  performance using root mean squared error, discrimination and calibration for 
  benefit. **Results:** Models including a linear interaction of the PI with
  treatment had adequate performance that was robust under most simulation 
  scenarios. Restricted cubic splines required larger sample sizes and higher 
  AUC of the PI to achieve adequate performance. The adaptive approach performed 
  equivalently to the best-performing method in each scenario. **Conclusion:** 
  Usually, models with just a linear interaction of the PI with treatment
  adequately predict absolute benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

In the simulated datasets based treatment was allocated at random using a 50:50
split. For each patient we simulated $8$ baseline covariates, where
$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Outcomes
for patients in the control arm were generated from a logistic regression model
including all baseline covariates. Coefficient values were such, so that the
prediction model had an AUC of $0.75$ and an event rate of $20\%$ in the control
arm was achieved. Under the base case scenario, outcomes in the treatment arm
were created using the same logistic regression model, including a constant
treatment odds ratio (OR) of $0.8$. The generated samples of the base case
scenario had sample size of 4,250 ($85%$ power for the detection of an
unadjusted OR of $0.8$).

We evaluated the effect of sample size considering additional scenarios with
sample sizes of `r 1064` and `r 17000`. We also evaluated the effect of prediction
performance, adjusting the baseline covariate coefficients, so that AUC values
of $0.65$ and $0.80$ were achieved when validating in a simulated dataset of
500,000 patients.

We simulated binary outcomes in the control arm using true probabilities 
$P(y=1|X) = \text{expit}(PI)$, where 
$\text{expit}(x)= \frac{e^x}{1 + e^x}$. 
In the treatment arm outcomes where generated with probabilities
$\text{expit}(lp_1)$ with
$$ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,$$
where the coefficients $\gamma_0, \gamma_1$ and $\gamma_2$ along with the
centering constant $c$ were set for each simulation scenario. In this way we
were able to assess a wide variety of scenarios, ranging from true constant
relative treatment effects to moderate and strong linear and quadratic
deviations. We also considered scenarios with treatment-covariate interactions.
These scenarios include 4 weak interactions
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
4 strong interactions 
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), 
and 2 weak and 2 strong interactions. Combining all these different settings
resulted in a simulation study of $66$ scenarios. The exact settings for each
scenario are available in the supplementary material.

@@

All methods assume that a risk prediction model is available and can be used to
assign individualized risk predictions. For the simulations we developed the
prediction models internally and blinded to treatment using logistic regression
including main effects for all baseline covariates and treatment. Risk
predictions on individuals were made setting treatment to $0$.

The *stratified HTE method* has been suggested as an alternative to
traditional subgroup analyses. Patients are stratified into equally-sized risk
strata---in this case based on risk quartiles. Absolute effects are estimated
using the differences in event rates between treatments within risk quarters. We
considered this approach as a reference, expecting it to perform worse than the
other candidates, as its objective is not individualized benefit prediction.

Another approach would be to assume *constant relative treatment effect* (OR) is
constant. In that case, absolute benefit is estimated from
$\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))$,
where $\text{expit}(x)=\frac{e^x}{1+e^x}$
and *lp* is the linear predictor of the prediction model.

A different approach fits a logistic regression using treatment, the prognostic
index and their interaction. In this case absolute benefit is estimated from
$\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)$.
We will refer to this method as the *linear interaction* approach.

Finally, we used *restricted cubic splines* (RCS) to relax the linearity assumption on
the effect of the linear predictor [@Harrell1988]. We compared the results
for 3, 4 and 5 knots when fitting the splines to introduce increasing
flexibility to the methods considered.

@@

For evaluating the prediction error of the considered methods we used root mean
squared error (RMSE), since both the true and the predicted benefits are known,
given that this is a simulation study. More specifically, we calculated RMSE
from
$$\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2$$

We also compared the discriminative ability of the methods under study. We
assessed discrimination using the c-for-benefit statistic described in
[@vanKlaveren2018]. Patients in each treatment arm are ranked based on their
predicted benefit and then are matched 1:1, dropping patients in the larger
treatment arm without a pair. We define *observed* pair-specific treatment
benefit as the difference of observed outcomes between the untreated and the
treated patient of each pair. Pair-specific *predicted* benefit is defined as
the average of predicted benefits within each pair. Then, c-for-benefit is
defined as the probability that from two randomly chosen matched patient pairs
with unequal observed benefit, the pair with greater observed benefit also has a
higher predicted benefit.

We evaluated calibration in a similar manner, using the integrated calibration
index (ICI) for benefit [@Austin2019]. After creating pairs based on predicted
benefit, observed benefits are regressed on the predicted benefits using a
locally weighted scatterplot smoother (loess). The ICI is the average absolute
difference between predicted and smooth observed benefit. Values closer to $0$
represent better calibration.

@@

The model including a constant relative treatment effect had the lowest median
RMSE in scenarios with a true constant relative treatment effect (OR = 0.8, N =
`r 4250` and AUC = 0.75) or moderate relative deviations were considered (Figure
\ref{fig:rmsebase}; Panel A). However, when we considered strong linear and
quadratic deviations from a constant relative treatment effect the linear
interaction model performed best (Figure \ref{fig:rmsebase}; Panels B and C).
Only in the case of strong quadratic deviations models including RCS (3 knots)
performed equally well to the linear interaction method. Increasing the number
of knots in RCS resulted in higher error rates across all scenarios. The
adaptive approach performed very similarly to the best performing model in each
scenario.

@@

When we increased the sample size (N = `r 17000`), the model including a
constant relative treatment effect had the lowest RMSE under the base case
scenario of constant relative treatment effects (Figure
\ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong linear
deviations the linear interaction model performed best (Figure
\ref{fig:rmsesamplesize}; Panels B and C). In the case of true strong quadratic
deviations from the base case of constant relative effects, the more flexible
RCS smoothing models (3 knots) had the lowest RMSE. Again, the increased
flexibility of RCS smoothing with higher number of knots resulted in overfitting
and worse performance (Figure \\ref{fig:rmsesamplesize}; Panel D).

### 2021-06-22 submission/manuscript.tex
reduce abstract word count

\title{Smooth risk-based predictive approaches to treatment effect
heterogeneity: A simulation study}

@@

\begin{abstract}
\textbf{Objective:} Compare different risk-based approaches to
estimating individualized treatment effects with simulations of the RCT
setting. \textbf{Study Design and Setting:} We compared different
methods for predicting absolute benefit on the basis of an available
prognostic index (PI) for risk: A model with the PI and a constant
relative treatment effect; models including an interaction of treatment
assignment with the PI; 4 quarters of the PI; nonlinear transformations
of the PI with increasing flexibility (restricted cubic splines with 3,
4 and 5 knots, respectively); an adaptive method, selecting the optimal
model based on Akaike's Information Criterion. Starting from a base case
scenario (sample size 4,250, average treatment effect odds ratio 0.8,
AUC of the PI 0.75), we considered diverse assumptions of the data
generation process of the ``true'' treatment effect, introducing linear
and quadratic interactions of the PI with treatment. We analyzed the
effect of varying sample size and discriminative ability of the PI. We
evaluated performance using root mean squared error, discrimination for
benefit and calibration for benefit. \textbf{Results:} The model
including a linear interaction of the risk linear predictor with
treatment had adequate performance that was robust under the majority of
the simulation scenarios. Methods using restricted cubic splines
required larger sample sizes and higher AUC of the PI to achieve
adequate performance. The adaptive approach's performance was comparable
to the performance of the best model in each scenario.
\textbf{Conclusion:} In most cases using a model just including a linear
interaction of the PI with treatment adequately predicts absolute
benefit.
\end{abstract}

\keywords{
    test
   \and
    test that
  }

@@

In the simulated datasets based treatment was allocated at random using
a 50/50 split. For each patient we simulated \(8\) baseline covariates,
where \(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
Outcomes for patients in the control arm were generated from a logistic
regression model including all baseline covariates. Coefficient values
were such, so that the prediction model had an AUC of \(0.75\) and an
event rate of \(20%
\) in the control arm was achieved. Under the base case scenario,
outcomes in the treatment arm were created using the same logistic
regression model, including a constant treatment odds ratio (OR) of
\(0.8\). The generated samples of the base case scenario were 4,250
(\(80\%\) power for the detection of an unadjusted OR of \(0.8\)).

We evaluated the effect of sample size considering additional scenarios
with sample sizes of 1,064 and 17,000. We also evaluated the effect of
prediction performance, adjusting the baseline covariate coefficients,
so that AUC values of \(0.65\) and \(0.80\) were achieved when
validating in a simulated dataset of \(500,000\) patients.

We simulated binary outcomes in the control arm based on
\(\text{expit}(PI)\), where \(\text{expit}(x)= \frac{e^x}{1 + e^x}\). In
the treatment arm outcomes where generated from \(\text{expit}(lp_1)\)
with \[ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,\] where the
coefficients \(\gamma_0, \gamma_1\) and \(\gamma_2\) along with the
centering constant \(c\) were set for each simulation scenario. In this
way we were able to assess a wide variety of scenarios, ranging from
true constant relative treatment effects to moderate and strong linear
and quadratic deviations. We also considered scenarios with
treatment-covariate interactions. These scenarios include 4 weak
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82\)), 4 strong
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61\)), and 2
weak and 2 strong interactions. Combining all these different settings
resulted in a simulation study of \(66\) scenarios. The exact settings
for each scenario are available in the supplementary material.

@@

All methods assume that a risk prediction model is available and can be
used to assign individualized predictions. For the simulations we
developed the prediction models internally and blinded to treatment
using logistic regression including main effects for all baseline
covariates and treatment. Predictions on individuals were made setting
treatment to \(0\).

The \textbf{stratified HTE method} was suggested as an alternative to
traditional subgroup analyses. Patients are stratified into
equally-sized risk strata---in this case based on risk quartiles.
Absolute effects are estimated using the differences in event rates
between treatments within risk quarters. We considered this approach as
a reference, expecting it to perform worse than the other candidates, as
its objective is not individualized benefit prediction.

We also considered a set of \textbf{linear methods}. We fit separate
models within treatment arms using only the treatment indicator and the
linear predictor of the internal risk prediction model. In the simpler
case, we assume a constant relative treatment effect (OR). Absolute
benefit is then estimated from \(\text{expit}(lp +\log(\text{OR}))\),
where \(\text{expit}(x)=\frac{e^x}{1+e^x}\) and \emph{lp} is the linear
predictor of the prediction model. A different approach fits a logistic
regression using treatment, risk linear predictor and their interaction.
In this case, absolute benefit is estimated from
\(\text{expit}(\beta_0+\beta_{lp}lp) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{lp}+\beta_*)lp)\).
We will refer to this method as the linear interaction approach.

Finally, we used restricted cubic splines (RCS) to relax the linearity
assumption on the effect of the linear predictor {[}7{]}. We compared
the results for 3, 4 and 5 knots when fitting the splines to introduce
increasing flexibility to the methods considered.

@@

Assuming that
\(\tau(\bm{x})=E\{y\:\vert\:\bm{x}, t_x=0\} - E\{y\:\vert\:\bm{x},t_x=1\}\)
is the true benefit for each patient and \(\hat{\tau}(x)\) is the
estimated benefit from a method under study, the ideal loss function to
use for the considered methods would be the unobservable root mean
squared error \(E\big\{(\hat{\tau} - \tau)^2\:\vert\:\bm{x}\big\}\).
However, in simulations the true patient-level benefit is available.
Therefore, we estimated RMSE from
\[\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2\]

We also compared the predictive performance of the methods under study.
We assessed discrimination using the c-for-benefit statistic described
in {[}8{]}. Patients in each treatment arm are ranked based on their
predicted benefit and then are matched 1:1, dropping patients in the
larger treatment arm without a pair. We define \emph{observed}
pair-specific treatment benefit as the difference of observed outcomes
between the untreated and the treated patient of each pair. Pair-specifi
\emph{predicted} benefit is defined as the average of predicted benefits
within each pair. Then, c-for-benefit is defined as the probability that
from two randomly chosen matched patient pairs with unequal observed
benefit, the pair with greater observed benefit also has a higher
predicted benefit.

We evaluated calibration in a similar manner, using the integrated
calibration index (ICI) for benefit {[}9{]}. After creating pairs based
on predicted benefit, observed benefits are regressed on the predicted
benefits using a locally weighted scatterplot smoother (loess). The ICI
is the area between the loess fit and the diagonal, with values closer
to \(0\) indicating better calibration.

@@

The model including a constant relative treatment effect had the lowest
median RMSE in scenarios where the base case of true constant relative
treatment effect (OR = 0.8, N = 4,250 and AUC = 0.75) or moderate
relative deviations were considered (Figure \ref{fig:rmsebase}; Panel
A). However, when we considered strong linear and quadratic deviations
from the base case the linear interaction model performed best (Figure
\ref{fig:rmsebase}; Panels B and C). Only in the case of strong
quadratic deviations models including RCS smoothing (3 knots) performed
equally well to the linear interaction method. Increasing the number of
knots in RCS smoothing resulted in higher error rates across all
scenarios. The performance of the adaptive approach was quite comparable
with the best performing model in each scenario.

@@

When we increased the sample size (N = 17,000), in all scenarios the
model more in agreement with the underlying settings had the lowest
error rates. Under the base case of constant relative treatment effects
(OR = 0.8), the model including a constant relative treatment effect had
the lowest RMSE (Figure \ref{fig:rmsesamplesize}; Panel A). When
introducing moderate and strong linear deviations the linear interaction
model performed best (Figure \ref{fig:rmsesamplesize}; Panels B and C).
In the case of true strong quadratic deviations from the base case of
constant relative effects, the more flexible RCS smoothing models (3
knots) had the lowest RMSE. Again, the increased flexibility of RCS
smoothing with higher number of knots resulted in overfitting and worse
performance (Figure \ref{fig:rmsesamplesize}; Panel D).

>>>>

\title{Continuous risk-based predictive approaches to treatment effect
heterogeneity: A simulation study}

@@

\begin{abstract}
\textbf{Objective:} Compare risk-based methods for individualizing
treatment effects with simulations in the RCT setting. \textbf{Study
Design and Setting:} We predicted absolute benefit based on an available
prognostic index (PI) using: a model with the PI and a constant relative
treatment effect; a model including an interaction of treatment with the
PI; 4 quarters of the PI; nonlinear transformations of the PI
(restricted cubic splines with 3, 4 and 5 knots); an adaptive model
selection method using Akaike's Information Criterion. Starting from a
base case scenario (sample size 4,250, constant odds ratio 0.8, AUC of
the PI 0.75), we considered diverse assumptions by introducing linear
and quadratic interactions of the PI with treatment and varying sample
size and discriminative ability of the PI. We evaluated performance
using root mean squared error, discrimination and calibration for
benefit. \textbf{Results:} Models including a linear interaction of the
PI with treatment had adequate performance that was robust under most
simulation scenarios. Restricted cubic splines required larger sample
sizes and higher AUC of the PI to achieve adequate performance. The
adaptive approach performed equivalently to the best-performing method
in each scenario. \textbf{Conclusion:} Usually, models with just a
linear interaction of the PI with treatment adequately predict absolute
benefit.
\end{abstract}

\keywords{
    treatment effect heterogeneity
   \and
    absolute benefit
   \and
    prediction models
  }

@@

In the simulated datasets based treatment was allocated at random using
a 50:50 split. For each patient we simulated \(8\) baseline covariates,
where \(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
Outcomes for patients in the control arm were generated from a logistic
regression model including all baseline covariates. Coefficient values
were such, so that the prediction model had an AUC of \(0.75\) and an
event rate of \(20\%\) in the control arm was achieved. Under the base
case scenario, outcomes in the treatment arm were created using the same
logistic regression model, including a constant treatment odds ratio
(OR) of \(0.8\). The generated samples of the base case scenario had
sample size of 4,250 (\(85%
\) power for the detection of an unadjusted OR of \(0.8\)).

We evaluated the effect of sample size considering additional scenarios
with sample sizes of 1,064 and 17,000. We also evaluated the effect of
prediction performance, adjusting the baseline covariate coefficients,
so that AUC values of \(0.65\) and \(0.80\) were achieved when
validating in a simulated dataset of 500,000 patients.

We simulated binary outcomes in the control arm using true probabilities
\(P(y=1|X) = \text{expit}(PI)\), where
\(\text{expit}(x)= \frac{e^x}{1 + e^x}\). In the treatment arm outcomes
where generated with probabilities \(\text{expit}(lp_1)\) with
\[ lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0,\] where the
coefficients \(\gamma_0, \gamma_1\) and \(\gamma_2\) along with the
centering constant \(c\) were set for each simulation scenario. In this
way we were able to assess a wide variety of scenarios, ranging from
true constant relative treatment effects to moderate and strong linear
and quadratic deviations. We also considered scenarios with
treatment-covariate interactions. These scenarios include 4 weak
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82\)), 4 strong
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61\)), and 2
weak and 2 strong interactions. Combining all these different settings
resulted in a simulation study of \(66\) scenarios. The exact settings
for each scenario are available in the supplementary material.

@@

All methods assume that a risk prediction model is available and can be
used to assign individualized risk predictions. For the simulations we
developed the prediction models internally and blinded to treatment
using logistic regression including main effects for all baseline
covariates and treatment. Risk predictions on individuals were made
setting treatment to \(0\).

The \emph{stratified HTE method} has been suggested as an alternative to
traditional subgroup analyses. Patients are stratified into
equally-sized risk strata---in this case based on risk quartiles.
Absolute effects are estimated using the differences in event rates
between treatments within risk quarters. We considered this approach as
a reference, expecting it to perform worse than the other candidates, as
its objective is not individualized benefit prediction.

Another approach would be to assume \emph{constant relative treatment
effect} (OR) is constant. In that case, absolute benefit is estimated
from \(\hat{\tau}(\bm{x}) = \text{expit}(PI +\log(\text{OR}))\), where
\(\text{expit}(x)=\frac{e^x}{1+e^x}\) and \emph{lp} is the linear
predictor of the prediction model.

A different approach fits a logistic regression using treatment, the
prognostic index and their interaction. In this case absolute benefit is
estimated from
\(\text{expit}(\beta_0+\beta_{PI}PI) - \text{expit}(\beta_0+\beta_{t_x}+(\beta_{PI}+\beta_*)PI)\).
We will refer to this method as the \emph{linear interaction} approach.

Finally, we used \emph{restricted cubic splines} (RCS) to relax the
linearity assumption on the effect of the linear predictor {[}7{]}. We
compared the results for 3, 4 and 5 knots when fitting the splines to
introduce increasing flexibility to the methods considered.

@@

For evaluating the prediction error of the considered methods we used
root mean squared error (RMSE), since both the true and the predicted
benefits are known, given that this is a simulation study. More
specifically, we calculated RMSE from
\[\text{RMSE}=\frac{1}{n}\sum_{i=1}^n\big(\tau(\bm{x}_i) - \hat{\tau}(\bm{x}_i)\big)^2\]

We also compared the discriminative ability of the methods under study.
We assessed discrimination using the c-for-benefit statistic described
in {[}8{]}. Patients in each treatment arm are ranked based on their
predicted benefit and then are matched 1:1, dropping patients in the
larger treatment arm without a pair. We define \emph{observed}
pair-specific treatment benefit as the difference of observed outcomes
between the untreated and the treated patient of each pair.
Pair-specific \emph{predicted} benefit is defined as the average of
predicted benefits within each pair. Then, c-for-benefit is defined as
the probability that from two randomly chosen matched patient pairs with
unequal observed benefit, the pair with greater observed benefit also
has a higher predicted benefit.

We evaluated calibration in a similar manner, using the integrated
calibration index (ICI) for benefit {[}9{]}. After creating pairs based
on predicted benefit, observed benefits are regressed on the predicted
benefits using a locally weighted scatterplot smoother (loess). The ICI
is the average absolute difference between predicted and smooth observed
benefit. Values closer to \(0\) represent better calibration.

@@

The model including a constant relative treatment effect had the lowest
median RMSE in scenarios with a true constant relative treatment effect
(OR = 0.8, N = 4,250 and AUC = 0.75) or moderate relative deviations
were considered (Figure \ref{fig:rmsebase}; Panel A). However, when we
considered strong linear and quadratic deviations from a constant
relative treatment effect the linear interaction model performed best
(Figure \ref{fig:rmsebase}; Panels B and C). Only in the case of strong
quadratic deviations models including RCS (3 knots) performed equally
well to the linear interaction method. Increasing the number of knots in
RCS resulted in higher error rates across all scenarios. The adaptive
approach performed very similarly to the best performing model in each
scenario.

@@

When we increased the sample size (N = 17,000), the model including a
constant relative treatment effect had the lowest RMSE under the base
case scenario of constant relative treatment effects (Figure
\ref{fig:rmsesamplesize}; Panel A). When introducing moderate and strong
linear deviations the linear interaction model performed best (Figure
\ref{fig:rmsesamplesize}; Panels B and C). In the case of true strong
quadratic deviations from the base case of constant relative effects,
the more flexible RCS smoothing models (3 knots) had the lowest RMSE.
Again, the increased flexibility of RCS smoothing with higher number of
knots resulted in overfitting and worse performance (Figure
\textbackslash ref\{fig:rmsesamplesize\}; Panel D).
